Syntax and Semantics of Abstract Binding Trees

### 2015-11-12 paper.tex
Sucky workaround for Euler lacking \varsigma, \varrho
[paper] consolidate the diagrams a bit more

@@

The interpretation of a variable
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P$ simply
projects it of the environment and embeds it into the model, as follows:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}} \arrow[r, "\IMode{\pi_3}"]
    \& {\IMode{V^{[\Gamma]}}} \arrow[r, "\IMode{\nu_{\Gamma}}"]
    \& {\IMode{P^{[\Gamma]}}} \arrow[r, "\IMode{\pi_x}"]
    \& \IMode{P_\tau}
  \end{tikzcd}
  \tag{var}
\]

Next interpret metavariable applications
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}
: \tau}_P$ where
$\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$, which
proceeds by projecting the metavariable's interpretation from the environment
and intantiating it via substitution:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
          \arrow[r, "\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}"]
    \&[4.0em] {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
          \arrow[r, "\IMode{\varsigma^\tau_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \&[1.0em] \IMode{P_\tau}
  \end{tikzcd}
  \tag{metavar}
\]
where
\vspace{-0.5em}
\begin{gather*}
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{S^{\{\Upsilon\}}}} \arrow[r, "\Define{\phi\,}{\,\Tuple{\pi_u}_{u\in\vec{u}}}"]
    \&[3.0em] {\IMode{S^{\{\vec{\sigma}\}}}}
  \end{tikzcd}
  \qquad\begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, "\Define{\psi\,}{\,\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}}"]
    \&[11.0em] {\IMode{P^{[\vec{\tau}]}}}
  \end{tikzcd}
\end{gather*}

The interpretation of operator applications is most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed by
symbol collections; therefore, operators must pass through suitable renamings in
order to be used in the interpretation. Let us begin by constructing for each
operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the morphism
$\SemBrackets{\vartheta}_P$ which shall rename the parameters of the operator
using the environment:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    |[alias = Src]|
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, swap, "\IMode{\pi_2}"]
    \&
    {\IMode{S^{\{\Upsilon\}}}}
      \arrow[r, swap, "\IMode{\FLift{\pi_{(-)}}{}}"]
    \&[1.5em]
    |[alias = Tgt]|
    {\IMode{\Operators_a}}
    \arrow[from = Src, to = Tgt, bend left, dashed, "\IMode{\SemBrackets{\vartheta}_{P}}"]
  \end{tikzcd}
\]

@@

The construction of $\beta$ proceeds by renaming the parameters of the operator
$\vartheta$ and constructing the (bound) exponentiated arguments $\gamma$ of the
operator.
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
       \arrow[r, "\Define{\beta\,}{\,\Tuple{\SemBrackets{\vartheta}_P, \lambda\gamma}}"]
    \&[4.0em] {\IMode{%
       \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}%
       \prod_{\MkValence{\vec{\sigma}_i}{\vec{\tau}_i}{\tau_i}\in\vec{v}}%
       P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}%
    }}
  \end{tikzcd}
\]
Argument $\lambda\IMode{{\gamma}_{i}}$ are the exponential transposes (curried
form) of the composites $\IMode{{\gamma}_{i}}$:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}}}
      \arrow[d, dashed, swap, "\IMode{\lambda{\gamma}_{i}}"]
    \&
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}
       \times\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}
    }
      \arrow[r, "\IMode{\Tuple{{\pi}_{[1,1]},{\phi}_{i},{\psi}_{i}}}"]
      \arrow[d, dashed, swap, "\IMode{{\gamma}_{i}}"]
    \&[3.5em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_{i}}_{P}}}
      \ar{dl}[
        , rotate = +19.25
        , xshift = -4.5em
        , yshift = -0.2em
      ]{\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_{i}\parallel\Gamma,\vec{\tau}_{i}\vdash M_i:\tau_i}_P}}
    \\[2.0em]
    {\IMode{{P}_{{\tau}_{i}}^{\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}}}
    \& {\IMode{P_{{\tau}_{i}}}}
    \&
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Pair{\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}}}{\,\Tuple{\pi_u\circ\pi_{[2,1]}}_{u\in\Dom{\vec{\sigma}_i}}}}"]
    \&[12.0em] {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}\\[-1.0em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Pair{\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}}}{\,\Tuple{\pi_x\circ\pi_{[2,1]}}_{x\in\Dom{\vec{\tau}_i}}}}"]
    \&[12.0em] {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}

>>>>

%% Euler doesn't have \varsigma and \varrho
\DeclareSymbolFont{greekletters}{OML}{cmr}{m}{it}
\DeclareMathSymbol{\varrho}{\mathalpha}{greekletters}{"25}
\DeclareMathSymbol{\varsigma}{\mathalpha}{greekletters}{"26}

@@

Variables are interpreted by the map
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P$ which
projects them from the environment and embeds them into the model. Metavariables
are resolved by the map
$\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}
  : \tau}_P$ (where
$\Lookup{\Omega}{\MV{m}}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$) which
projects their interpretation from the environment and intantiates it via
substitution:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{P_\tau}}
    \&[8.0em] {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[rr, dashed, "\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x:\tau}_P"]
      \ar{drr}[
        , rotate = -16.5
        , xshift = +1.5em
        , yshift = -0.2em
        , swap
      ]{{\nu}_{\Gamma} {\pi}_{3}}
      \ar{dl}[
        , rotate = +11.5
        , xshift = -2.75em
      ]{\IMode{\Tuple{\pi_\MV{m}\,\pi_1, \phi\,\pi_2, \psi}}}
      \arrow[l, dashed, swap, "\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash\MApp{\MV{m}}{\vec{u}}{\vec{M}}: \tau}_P"]
    \&[-2.0em]
    \&[+3.0em]{\IMode{P_\tau}}
    \\
    {\IMode{P_{\tau}^{\Yoneda{\vec{\sigma}\parallel\vec{\tau}}}\times S^{\{\vec{\sigma}\}} \times P^{[\vec{\tau}]}}}
      \arrow[u, "\IMode{\varsigma^\tau_{\vec{\sigma}\parallel\vec{\tau}}}"]
    \&{}
    \&{}
    \& {\IMode{P^{[\Gamma]}}}
      \arrow[u, swap, "{\pi}_{x}"]
    \\
    {\IMode{P^{[\vec{\tau}]}}}
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[l, swap, "\Define{\psi\,}{\,\Tuple{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash M:\tau}_P}_{(M,\tau)\in (\vec{M},\vec{\tau})}}"]
    \& {\IMode{S^{\{\Upsilon\}}}}
      \arrow[r, "\Define{\phi\,}{\,\Tuple{\pi_u}_{u\in\vec{u}}}"]
    \& {\IMode{S^{\{\vec{\sigma}\}}}}
  \end{tikzcd}
\]

Interpretation of operator applications is the most complicated. Recall that,
unlike in standard treatments of universal algebra, our operators are indexed by
symbol collections; therefore, operators must pass through suitable renamings in
order to be used in the interpretation. Let us begin by constructing for each
operator $\IsOperator{\Upsilon}{\vartheta}{a}$ the morphism
$\SemBrackets{\vartheta}_P$ which shall rename the parameters of the operator
using the environment:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    |[alias = Src]|
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
      \arrow[r, swap, "\IMode{\pi_2}"]
    \&
    {\IMode{S^{\{\Upsilon\}}}}
      \arrow[r, swap, "\IMode{\FLift{\pi_{(-)}}{}}"]
    \&[1.5em]
    |[alias = Tgt]|
    {\IMode{\Operators_a}}
    \arrow[from = Src, to = Tgt, bend left, dashed, "\IMode{\SemBrackets{\vartheta}_{P}}"]
  \end{tikzcd}
\]

@@

The construction of $\beta$ proceeds by renaming the parameters of the operator
$\vartheta$ and constructing the (bound) exponentiated arguments $\gamma$ of the
operator.
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
       \arrow[r, "\Define{\beta\,}{\,\Tuple{\SemBrackets{\vartheta}_P, \lambda\gamma}}"]
    \&[4.0em] {\IMode{%
       \coprod_{\vartheta\in \Operators_{\MkArity{\vec{v}}{\tau}}}%
       \prod_{\MkValence{\vec{\sigma}_i}{\vec{\tau}_i}{\tau_i}\in\vec{v}}%
       P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}%
    }}
  \end{tikzcd}
\]
Arguments $\lambda\IMode{{\gamma}_{i}}$ are the exponential transposes (curried
form) of the composites $\IMode{{\gamma}_{i}}$:
\[
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}}}
      \arrow[d, dashed, swap, "\IMode{\lambda{\gamma}_{i}}"]
    \&
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_{P}
       \times\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}
    }
      \arrow[r, "\IMode{\Tuple{{\pi}_{[1,1]},{\phi}_{i},{\psi}_{i}}}"]
      \arrow[d, dashed, swap, "\IMode{{\gamma}_{i}}"]
    \&[3.5em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_{i}}_{P}}}
      \ar{dl}[
        , rotate = +19.25
        , xshift = -4.5em
        , yshift = -0.2em
      ]{\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_{i}\parallel\Gamma,\vec{\tau}_{i}\vdash M_i:\tau_i}_P}}
    \\[2.0em]
    {\IMode{{P}_{{\tau}_{i}}^{\Yoneda{\vec{\sigma}_{i}\parallel\vec{\tau}_{i}}}}}
    \& {\IMode{P_{{\tau}_{i}}}}
    \&
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[cramped, ampersand replacement = \&]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Pair{\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}}}{\,\Tuple{\pi_u\circ\pi_{[2,1]}}_{u\in\Dom{\vec{\sigma}_i}}}}"]
    \&[12.0em] {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}\\[-1.0em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Pair{\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}}}{\,\Tuple{\pi_x\circ\pi_{[2,1]}}_{x\in\Dom{\vec{\tau}_i}}}}"]
    \&[12.0em] {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}
